---
title: "marinemammalstrandings"
author: "Zoe Zilz"
date: "6/29/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("janitor")
library(tidyverse)
library(janitor)
library(lubridate)
library(scales)
library(stringr)
```

# Summary of Marine Mammal Stranding Data from 1983-2020


### Importing Data
Want to keep all rows with mammals that are DOA (not "alive"), spatial info, date, locality + detail, left/euthanized/died at site?,
```{r data wrangling}
mammals20 <- read_csv("MMstranding_2006_2020.csv") %>%
  clean_names() %>% 
  select(common_name, city, locality_detail, observation_date, how_observed, observation_status, date_of_examination, condition_at_examination, left_at_site, euthanized_at_site, died_at_site, relocated, condition_comments, sex, age_class, whole_carcass, partial_carcass, length, length_units, weight, weight_units, carcass_left_at_site, carcass_buried, carcass_other, carcass_other_description, additional_remarks) # we want these cols
  #filter(observation_status != "Alive"| euthanized_at_site == "Y" | died_at_site == "Y")  # only keep dead stuff

dates <- mdy(mammals20$observation_date) #make sure dates are dates

years <- year(dates)

mammals20 <- mammals20 %>% 
  mutate(year_stranded = years)

############################################### up to 2005 data #####################################  

mammals05 <- read_csv("MMstranding_1983_2005.csv") %>% 
  clean_names() %>%  #makes col headers easier to work with -- no spaces
  select(common_name, date_of_occurance, city, location_details, type_of_occurance, initial_condition, sex, age, length, length_units, weight, weight_units, final_disposition_remarks) %>% 
  rename(observation_date = date_of_occurance, 
         locality_detail = location_details,
         observation_status = initial_condition,
         age_class = age,
         additional_remarks = final_disposition_remarks)
  #filter(observation_status != "ALIVE") # do this after combo

dates05 <- mdy(mammals05$observation_date) #make sure dates are dates

years05 <- year(dates05)

mammals05 <- mammals05 %>% 
  mutate(year_stranded = years05)

############################################# COMBINING THEM ######################################

combo20 <- mammals20 %>% 
  select(common_name, observation_date, city, locality_detail, observation_status, sex, age_class, length, length_units, weight, weight_units, additional_remarks, year_stranded, carcass_left_at_site)

combo05 <- mammals05 %>% 
  select(common_name, observation_date, city, locality_detail, observation_status, sex, age_class, length, length_units, weight, weight_units, additional_remarks, year_stranded) %>% 
  mutate(carcass_left_at_site = NA)

live_mammals <- rbind(combo20, combo05) %>%   # need to tidy this up so all of the species names are the same
  mutate(sex2 = case_when(
    sex == "MALE" ~ "M",
    sex == "FEMALE" ~ "F",
    sex == "UNKNOWN" ~ "U",
    TRUE ~ as.character(sex)
  )) %>% 
  mutate(sex2 = replace_na(sex2, "UNKNOWN")) %>%
  mutate(age_class2 = case_when(
    age_class == "PUP"|age_class == "CALF" ~ "PUP/CALF",
    TRUE ~ as.character(age_class)
  )) %>% 
  mutate(age_class2 = replace_na(age_class2, "UNKNOWN")) %>% # is this the only way to get rid of NAs?
  mutate(common_name2 = case_when(
    common_name == "Cetacean, unidentified" ~ "UNIDENTIFIED CETACEAN",
    common_name == "Dolphin, bottlenose" ~"BOTTLENOSE DOLPHIN"  ,
    common_name == "Dolphin, common, long-beaked" ~ "LONG BEAKED COMMON DOLPHIN" ,
    common_name == "Dolphin, Risso's" ~ "RISSO'S DOLPHIN" ,
    common_name == "Dolphin, northern right whale" ~ "NORTHERN RIGHT WHALE",
    common_name == "Dolphin, striped" ~ "STRIPED DOLPHIN" ,
    common_name == "Dolphin, common, short-beaked" ~ "SHORT BEAKED COMMON DOLPHIN" ,
    common_name == "SHORT-BEAKED COMMON DOLPHIN" ~ "SHORT BEAKED COMMON DOLPHIN" ,
    common_name == "LONG-BEAKED COMMON DOLPHIN" ~ "LONG BEAKED COMMON DOLPHIN" ,
    common_name == "COMMON DOLPHIN" ~ "UNIDENTIFIED COMMON DOLPHIN" ,
    common_name == "Dolphin, Pacific white-sided" ~"PACIFIC WHITE-SIDED DOLPHIN" ,
    common_name == "Dolphin, unidentified" ~ "UNIDENTIFIED DOLPHIN",
    common_name == "Marine mammal, unidentified" ~ "UNIDENTIFIED",
    common_name == "Otter, sea" ~ "SEA OTTER",
    common_name == "pinniped, unidentified"
    |common_name =="Pinniped, unidentified"
    |common_name =="unidentified pinniped"
    |common_name == "Seal, unidentified"  ~ "UNIDENTIFIED PINNIPED",
    common_name == "Porpoise, Dall's" ~"DALL'S PORPOISE"   ,
    common_name == "Porpoise, harbor" ~ "HARBOR PORPOISE"  ,
    common_name == "Porpoise, unidentified phocoenidae" ~ "UNIDENTIFIED PORPOISE",
    common_name == "Sea lion, California"  ~ "CALIFORNIA SEA LION" ,
    common_name == "Sea lion, Steller" ~ "STELLAR SEA LION",
    common_name == "Sea lion, unidentified" ~ "UNIDENTIFIED SEA LION",
    common_name == "Seal, Northern fur" ~ "NORTHERN FUR SEAL"  ,
    common_name == "Seal, Guadalupe fur"~ "GUADALUPE FUR SEAL",
    common_name == "Seal, northern elephant"  ~"NORTHERN ELEPHANT SEAL"  ,
    common_name == "Seal, harbor" ~ "HARBOR SEAL"   ,
    common_name == "Whale, sperm" ~ "SPERM WHALE" ,
    common_name == "Whale, blue" ~ "BLUE WHALE",
    common_name == "Whale, minke" ~ "MINKE WHALE",
    common_name == "Whale, fin"  ~ "FIN WHALE"  ,
    common_name == "Whale, Cuvier's beaked" ~ "CUVIER'S BEAKED WHALE"  ,
    common_name == "Whale, unidentified beaked"
    |common_name == "Whale, unidentified baleen"  ~ "UNIDENTIFIED WHALE",
    common_name == "Whale, gray"  ~"GRAY WHALE"  ,
    common_name == "Whale, humpback" ~ "HUMPBACK WHALE",
    common_name == "Whale, pygmy sperm" ~ "PYGMY SPERM WHALE" ,
    common_name == "Whale, dwarf sperm" ~ "DWARF SPERM WHALE",
    TRUE ~ as.character(common_name)  # weird syntax for "otherwise, do nothing"
  ))

mammals <- live_mammals %>% 
  filter(observation_status != "ALIVE") #ugh, i'll just have to eliminate the died or euthanized on site, they were all likely removed anyway?
```

##How many mammals stranded per year?
```{r yearly strandings}
yearly_strandings <- as.data.frame(table(mammals$year_stranded)) %>% 
  #pivot_wider() %>% # not sure if this is necessary?
  rename(year = Var1,
         total_dead = Freq) # new name = old name syntax

barp <- ggplot(yearly_strandings)+
  geom_col(aes(x = year, y = total_dead, fill = I("dodgerblue4")))+ # geom col represents values
  scale_x_discrete(breaks = pretty_breaks())+
  ylab("Marine Mammal Carcasses Count")+
  xlab("Year")+
  geom_vline(aes(xintercept = 2005, colour = I("red")))+ # add line for 2005 when data collection method changed
  theme_classic()

png("annual_mm_count.png", height = 300, width = 450, units = "px")
barp
dev.off()

mean_yrly_dead <- mean(yearly_strandings$total_dead)
sd_yrly_dead <- sd(yearly_strandings$total_dead)
```


##What is the biomass of marine mammal carcasses?
- lots of blanks for weight, so we need to fill those in with means
- different mean for each size class
- general mean if size class is also unknown

###We want to know how much is washed ashore.
```{r wrangling weight data}
# first have to convert all the lb weights into kilograms
convert_pounds <- 0.453592
# if weight units is lbs, multiply weight by convert_pounds; otherwise, do nothing
live_mammals$weight_kg <- live_mammals$weight * ifelse(live_mammals$weight_units=="lb",convert_pounds,1)

# we can apparently also calculate weight from this length relationship?

# now that everything is in kg, we can get mean weights for pup/calf, pup, calf, yearling, subadult, adult
# we have more data on weights if we use the live + dead dataset
mean_weights <- live_mammals %>% 
  group_by(common_name2, age_class2, sex2) %>% 
  summarise(
    mean_wt = mean(weight_kg, na.rm=TRUE), 
    sd = sd(weight_kg, na.rm=TRUE),
    n = n())



#write_csv(mean_weights, "marinemammalweights.csv") # there are so many unknows I just need to export and manually fill in from the internet, only doing this once

```

Need to do some extrapolation to fill in blanks...
https://stackoverflow.com/questions/38687674/replace-na-with-mean-based-on-row-subset-matching-another-column
^ not helpful becayse I need to manyally fill in some
```{r extrapolating biomass to unknowns based on averages}
meanweights <- read_csv("marinemammalweights.csv") %>% 
  mutate(id = paste(common_name2, age_class2, sex2, sep = "_")) # create matchable ID

biomass <- mammals %>% 
  mutate(common_name2 = case_when(common_name2 == "UNIDENTIFIED CETACEAN" ~ "UNIDENTIFIED WHALE",
                                  common_name2 == "unidentified mysticete" ~ "UNIDENTIFIED WHALE",
                                   TRUE ~ as.character(common_name2))) %>%  # had to remove this redundant bit ebcause there is no matching "cetacean" in averages df
  mutate(sex2 = case_when(sex2 == "UNKNOWN" ~ "U",
                                   TRUE ~ as.character(sex2))) %>% 
  mutate(id = paste(common_name2, age_class2, sex2, sep = "_"))  # create matchable ID

biomass$weight_kg <- biomass$weight * ifelse(biomass$weight_units=="lb",convert_pounds,1) # add weight in kg

#trying with data_table
setDT(biomass)
setDT(meanweights)

biomass[is.na(weight_kg), # in biomass, if weight_kg = NA, then:
        weight_kg := meanweights # edit rows in weight_kg with values from meanweights
        [.SD, # but only rows in the .SD(subset data, so rows w/NAs)
          on=.(id), #and only if they match on the id column (which can be multiple columns)
          x.mean_wt]] #replace with values from meanweights$mean_wt (x. means in previously stated datasex)
# syntax:
# DF[i, j] = biomass[if this, do this]
# in j, .SD refers to the subset of data created by if-filtering done in i
# when there is a := in j, this modifies columns in the datatable
# using .SD to look up rows in meanweights on id and fetching values in column mean_wt, and modifying rows in biomass using :=
# https://stackoverflow.com/questions/50010196/replacing-na-values-from-another-dataframe-by-id
# https://stackoverflow.com/questions/49984172/comparing-values-in-multiple-columns-in-r-dataframes-updating-missing-values/49984344#49984344

# checking everything filled in:
# leave out na.rm = TRUE so that it throws an NA if tehre is *even one* NA
checkcheck <- biomass %>% 
  group_by(id) %>% 
  summarise(kg = sum(weight_kg))

```

What is the yearly input?
```{r mammal dead biomss graph}
# get sum of biomass per year
annual_biomass_bysp <- biomass %>% 
  group_by(year_stranded, common_name2) %>% 
  summarise(kg = sum(weight_kg))

annual_biomass <- biomass %>% 
  group_by(year_stranded) %>% 
  summarise(kg = sum(weight_kg))

# without whales
annual_nowhales <- biomass %>% 
  filter(!str_detect(common_name2, 'WHALE'))%>%  #stringr detect
  group_by(year_stranded) %>% 
  summarise(kg = sum(weight_kg))

# just for visualization
#ggplot(annual_biomass)+
#  #geom_line(aes(x = year_stranded, y = kg, colour = common_name2))+ # by species -- horrible mess it turns out
#  geom_line(aes(x = year_stranded, y = kg))+ # sum total
#  geom_line(data = annual_nowhales, aes(x = year_stranded, y = kg, colour = I("dodgerblue4")))+
#  scale_y_continuous(breaks = pretty_breaks())+ # force integers on y axis
#  ylab("Kilograms of Dead Marine Mammal Biomass")+
#  xlab("Year")+
#  theme_classic()

nowhales <- ggplot(annual_biomass)+
  geom_line(data = annual_nowhales, aes(x = year_stranded, y = kg, colour = I("dodgerblue4")))+
  scale_y_continuous(breaks = pretty_breaks())+ # force integers on y axis
  ylab("Biomass (kg)")+
  xlab("Year")+
  geom_vline(aes(xintercept = 2005, colour = I("red")))+ # add line for 2005 when data collection method changed
  theme_classic()

total <- ggplot(annual_biomass)+
  #geom_line(aes(x = year_stranded, y = kg, colour = common_name2))+ # by species -- horrible mess it turns out
  geom_line(aes(x = year_stranded, y = kg, colour = I("dodgerblue4")))+ # sum total
  scale_y_continuous(breaks = pretty_breaks())+ # force integers on y axis
  ylab("Biomass (kg)")+
  xlab("Year")+
  geom_vline(aes(xintercept = 2005, colour = I("red")))+ # add line for 2005 when data collection method changed
  theme_classic()

png("annual_mm_biomass.png", height = 300, width = 450, units = "px")
total
dev.off()

png("annual_mm_biomass_nowhales.png", height = 300, width = 450, units = "px")
nowhales
dev.off()
```

###And how much is left to be eaten.
```{r biomass}
carcasses <- biomass %>% 
  filter(carcass_left_at_site == "Y")%>%  #only keep if carcass stayed on beach 
  group_by(year_stranded) %>% 
  summarise(kg = sum(weight_kg))

carcass <- ggplot(carcasses)+
  geom_line(aes(x = year_stranded, y = kg, colour = I("dodgerblue4")))+ # sum total
  scale_y_continuous(breaks = pretty_breaks())+ # force integers on y axis
  scale_x_continuous(breaks = seq(2006, 2020, by = 1))+
  ylab("Biomass (kg)")+
  xlab("Year")+
  theme_classic()

png("annual_carcass_biomass.png", height = 300, width = 450, units = "px")
carcass
dev.off()

mean_carcass_biomass <- mean(carcasses$kg)
```

